---
title: "Running MCEM in NIMBLE"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The Monte Carlo expectation maximization (MCEM) algorithm is a method for finding maximum likelihoods in models with latent or missing information. MCEM uses Monte Carlo integration methods to simplify and evaluate potentially high-dimensional integrals.

Here we use the pump model example from BUGS to illustrate how to implement the MCEM algorithm to find maximum likelihood estimates of parameters for statistical models with latent states.

```{r}
library(nimble, warn.conflicts = FALSE)
```

#### NIMBLE code for the pump example 

```{r}
pumpCode <- nimbleCode({ 
  for (i in 1:N){
    theta[i] ~ dgamma(alpha, beta)
    lambda[i] <- theta[i] * t[i]
    x[i] ~ dpois(lambda[i])
  }
  alpha ~ dexp(1.0)
  beta ~ dgamma(0.1, 1.0)
})
```

#### Set up the model

Assign values to the constants in the model, provide data, initialize the random variables, and then create the model.

```{r}
pumpConsts <- list(N = 10,
                   t = c(94.3, 15.7, 62.9, 126, 5.24,
                         31.4, 1.05, 1.05, 2.1, 10.5))

pumpData <- list(x = c(5, 1, 5, 14, 3, 19, 1, 1, 4, 22))

pumpInits <- list(alpha = 0.1, beta = 0.1,
                  theta = rep(0.1, pumpConsts$N))

## Create the model
pump <- nimbleModel(code = pumpCode, name = "pump", constants = pumpConsts,
                    data = pumpData, inits = pumpInits, buildDerivs = TRUE)

```

#### Compile the model

```{r}
Cpump <- compileNimble(pump)
```

#### Run the MCEM algorithm provided by NIMBLE

```{r}
pumpMCEM <- buildMCEM(model = pump, latentNodes = "theta[1:10]")
CpumpMCEM <- compileNimble(pumpMCEM, project = pump)
pumpMLE <- CpumpMCEM$findMLE()
pumpMLE
```
