# Posterior predictive sampling

Once one has samples from an MCMC, one often wants to do some post hoc manipulation of the samples. A important example is posterior predictive sampling, which is needed for posterior predictive checking.

With posterior predictive sampling, we need to simulate new data values, once for each posterior sample. These samples can then be compared with the actual data as a model check.

In this example, we'll follow the posterior predictive checking done in the Gelman et al. Bayesian Data Analysis book, using Newcomb's speed of light measurements.

## Posterior predictive sampling using a loop in R

Simon Newcomb made 66 measurements of the speed of light, which one might model using a normal distribution. One question discussed in Gelman et al. is whether the lowest measurements, which look like outliers, could have reasonably come from a normal distribution.

### Setup

We set up the nimble model. 

```{r}
library(nimble, warn.conflicts = FALSE)

code <- nimbleCode({
    ## noninformative priors
    mu ~ dflat()
    sigma ~ dhalfflat()
    ## likelihood
    for(i in 1:n) {
     	y[i] ~ dnorm(mu, sd = sigma)
    }
})

data <- list(y = MASS::newcomb)
inits <- list(mu = 0, sigma = 5)
constants <- list(n = length(data$y))

model <- nimbleModel(code = code, data = data, constants = constants, inits = inits)

## Ensure we have the nodes needed to simulate new datasets
dataNodes <- model$getNodeNames(dataOnly = TRUE)
parentNodes <- model$getParents(dataNodes, stochOnly = TRUE)  # `getParents` is new in nimble 0.11.0
## Ensure we have both data nodes and deterministic intermediates (e.g., lifted nodes)
simNodes <- model$getDependencies(parentNodes, self = FALSE)

cmodel 	<- compileNimble(model)
mcmc    <- buildMCMC(model, monitors = parentNodes)
cmcmc   <- compileNimble(mcmc, project = model)

samples <- runMCMC(cmcmc, niter = 1000, nburnin = 500)
```

### Posterior predictive sampling by direct variable assignment

We'll loop over the samples and use the compiled model (uncompiled would be ok too, but slower) to simulate new datasets.

```{r}
nSamp <- nrow(samples)
n <- length(data$y)
ppSamples <- matrix(0, nSamp, n)

set.seed(1)
for(i in 1:nSamp){
  cmodel[["mu"]] <- samples[i, "mu"]             ## or cmodel$mu <- samples[i, "mu"]
  cmodel[["sigma"]] <- samples[i, "sigma"]
  cmodel$simulate(simNodes, includeData = TRUE)
  ppSamples[i, ] <- cmodel[["y"]]
}
```

### Posterior predictive sampling using `values`

That's fine, but we needed to manually insert values for the different variables. For a more general solution, we can use nimble's `values` function as follows.

```{r}
ppSamples <- matrix(0, nrow = nSamp, ncol =
          length(model$expandNodeNames(dataNodes, returnScalarComponents = TRUE)))

## Determine ordering of variables in `mvSamples` modelValues and therefore in `samples`:
vars <- cmcmc$mvSamples$getVarNames()

## Quick check of variable ordering so we know `values()` is given things in the right order.
## This should generally be the case, but best to check.
vars
colnames(samples)

set.seed(1)
system.time({
for(i in seq_len(nSamp)) {
    values(cmodel, vars) <- samples[i, ]  # assign 'flattened' values
    cmodel$simulate(simNodes, includeData = TRUE)
    ppSamples[i, ] <- values(cmodel, dataNodes)
}
})
```

### Doing the posterior predictive check

At this point, we can implement the check we want using our chosen discrepancy measure. Here a simple check uses the minimum observation.

```{r}
obsMin <- min(data$y)
ppMin <- apply(ppSamples, 1, min)

# ## Check with plot in Gelman et al. (3rd edition), Figure 6.3
hist(ppMin, xlim = c(-50, 20),
    main = "Discrepancy = min(y)", 
    xlab = "min(y_rep)")
abline(v = obsMin, col = 'red')
```

## Posterior predictive sampling using a nimbleFunction

The approach above could be slow, even with a compiled model, because the loop is carried out in R. We could instead do all the work in a compiled nimbleFunction.

### Writing the nimbleFunction

Let's set up a nimbleFunction. In the setup code, we'll manipulate the nodes and variables, similarly to the code above. In the run code, we'll loop through the samples and simulate, also similarly.

Remember that all querying of the model structure needs to happen in the setup code. We also need to pass in the MCMC object so that we can determine at setup time the names of the variables we are copying from the posterior samples into the model.

The run code takes the actual samples as the input argument, so the nimbleFunction will work regardless of how long the MCMC was run for.

```{r}
ppSamplerNF <- nimbleFunction(
          setup = function(model, mcmc) {
              dataNodes <- model$getNodeNames(dataOnly = TRUE)
              parentNodes <- model$getParents(dataNodes, stochOnly = TRUE)
              cat("Stochastic parents of data are:", paste(parentNodes, collapse = ','), ".\n")
              simNodes <- model$getDependencies(parentNodes, self = FALSE)
              vars <- mcmc$mvSamples$getVarNames()  # need ordering of variables in mvSamples / samples matrix
              cat("Using posterior samples of:", paste(vars, collapse = ','), ".\n")
              n <- length(model$expandNodeNames(dataNodes, returnScalarComponents = TRUE))
          },
          run = function(samples = double(2)) {
              nSamp <- dim(samples)[1]
              ppSamples <- matrix(nrow = nSamp, ncol = n)   
              for(i in 1:nSamp) {
                    values(model, vars) <<- samples[i, ]
                    model$simulate(simNodes, includeData = TRUE)
                    ppSamples[i, ] <- values(model, dataNodes)
              }
              returnType(double(2))       
              return(ppSamples)
          })
```

### Using the nimbleFunction  

```{r}
colnames(samples)  ## again, to check ordering of vars is same in 'vars' and in 'samples'

## Create the sampler for this model and this MCMC.
ppSampler <- ppSamplerNF(model, mcmc)
cppSampler <- compileNimble(ppSampler, project = model)

set.seed(1)
system.time(ppSamples_via_nf <- cppSampler$run(samples))

identical(ppSamples, ppSamples_via_nf)
```

So we get exactly the same results (note the use of `set.seed`) but much faster.

Here the speed doesn't really matter but for more samples and larger models it often will, even after accounting for the time spent to compile the nimbleFunction.

