# MCMC for logistic regression with random effects

This example shows how to build and run MCMC for a generalized linear mixed model (GLMM), specifically a logistic regression model with random effects.

## Model Creation

```{r}
## load the NIMBLE library
library(nimble)

## define the model
code <- nimbleCode({
    beta0 ~ dnorm(0, sd = 10000)
    beta1 ~ dnorm(0, sd = 10000)
    sigma_RE ~ dunif(0, 1000)
    for(i in 1:N) {
        beta2[i] ~ dnorm(0, sd = sigma_RE)
        logit(p[i]) <- beta0 + beta1 * x[i] + beta2[i]
        r[i] ~ dbin(p[i], n[i])
    }
})

## constants, data, and initial values
constants <- list(N = 10)

data <- list(
    r = c(10, 23, 23, 26, 17, 5, 53, 55, 32, 46),
    n = c(39, 62, 81, 51, 39, 6, 74, 72, 51, 79),
    x = c(0,  0,  0,  0,  0,  1, 1,  1,  1,  1)
)

inits <- list(beta0 = 0, beta1 = 0, sigma_RE = 1)

## create the model object
Rmodel <- nimbleModel(code=code, constants=constants, data=data, inits=inits, check = FALSE)
```

## Default MCMC Algorithm

Now we are ready to create the default MCMC algorithm from model object
```{r}
Rmcmc <- buildMCMC(Rmodel)
```

## Compile the model and MCMC algorithm
```{r}
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
```

## Execute MCMC algorithm and extract samples
```{r}
Cmcmc$run(10000)
samples <- as.matrix(Cmcmc$mvSamples)
```

## Customize an MCMC Algorithm

First we make a new instance of the model:
```{r}
Rmodel <- nimbleModel(code=code, constants=constants, data=data, inits=inits)
```

Then we make an empty MCMC configuration and add some samplers of our choice to it:
```{r}
spec <- configureMCMC(Rmodel, nodes=NULL)
spec$addSampler(type = 'slice', target ='beta0')
spec$addSampler(type = 'slice', target ='beta1')
spec$addSampler(type = 'RW', target ='sigma_RE')
spec$addSampler(type = 'RW_block', target ='beta2[1:10]')
```

Then we build the MCMC from the configuration:
```{r}
Rmcmc <- buildMCMC(spec)
```


## Compile model and custom MCMC algorithm
```{r}
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
```

## Execute custom MCMC algorithm and extract samples
```{r}
Cmcmc$run(10000)
samples <- as.matrix(Cmcmc$mvSamples)
```