# MCMC for logistic regression with random effects

This example shows how to build and run MCMC for a generalized linear mixed model (GLMM), specifically a logistic regression model with random effects.

## Model Creation

```{r}
## load the NIMBLE library
library(nimble)

## define the model
code <- nimbleCode({
    beta0 ~ dnorm(0, sd = 10000)
    beta1 ~ dnorm(0, sd = 10000)
    sigma_RE ~ dunif(0, 1000)
    for (i in 1:N) {
        beta2[i] ~ dnorm(0, sd = sigma_RE)
        logit(p[i]) <- beta0 + beta1 * x[i] + beta2[i]
        r[i] ~ dbin(p[i], n[i])
    }
})

## constants, data, and initial values
constants <- list(N = 10)

data <- list(
    r = c(10, 23, 23, 26, 17, 5, 53, 55, 32, 46),
    n = c(39, 62, 81, 51, 39, 6, 74, 72, 51, 79),
    x = c(0,  0,  0,  0,  0,  1, 1,  1,  1,  1)
)

inits <- list(beta0 = 0, beta1 = 0, sigma_RE = 1)

## create the model object
glmmModel <- nimbleModel(code = code, constants = constants, data = data, 
                         inits = inits, check = FALSE)
```

## Default MCMC Algorithm

Now we are ready to create the default MCMC algorithm from model object.

```{r}
glmmMCMC <- buildMCMC(glmmModel)
```

## Compile the model and MCMC algorithm

```{r}
CglmmModel <- compileNimble(glmmModel)
CglmmMCMC <- compileNimble(glmmMCMC, project = glmmModel)
```

## Execute MCMC algorithm and extract samples
```{r}
samples <- runMCMC(CglmmMCMC, niter = 10000)
```

## Customize an MCMC Algorithm

We may want to customize the MCMC with specific samplers. To do so, we first make a new instance of the model:

```{r}
glmmModel <- nimbleModel(code = code, constants = constants, 
                         data = data, inits = inits)
```

Then we make an empty MCMC configuration and add some samplers of our choice to it:

```{r}
spec <- configureMCMC(glmmModel, nodes = NULL)
spec$addSampler(type = 'slice', target = 'beta0')
spec$addSampler(type = 'slice', target = 'beta1')
spec$addSampler(type = 'RW', target = 'sigma_RE')
spec$addSampler(type = 'RW_block', target = 'beta2[1:10]')
```

Then we build the MCMC from the configuration:

```{r}
customGlmmMCMC <- buildMCMC(spec)
```


## Compile model and custom MCMC algorithm
```{r}
CglmmModel <- compileNimble(glmmModel)
CglmmMCMC <- compileNimble(customGlmmMCMC, project = glmmModel)
```

## Execute custom MCMC algorithm and extract samples
```{r}
samples <- runMCMC(CglmmMCMC, niter = 10000)
```

Note that NIMBLE does not provide its own tools for working with posterior samples.
