# Building a model from BUGS code in R using NIMBLE

The BUGS language for declaring statistical models was popularized by [WinBUGS](http://www.mrc-bsu.cam.ac.uk/software/bugs/), [OpenBUGS](http://www.openbugs.net/w/FrontPage) and [JAGS](http://mcmc-jags.sourceforge.net/).  Those generate and run an MCMC for the model, but they don't allow a programmer to use the model in any other way.  NIMBLE provides a new implementation of the BUGS language and creates model objects you can program with.

NIMBLE supports most models written for BUGS or JAGS.  It also extends the BUGS language in a bunch of ways that we won't go into here. See the [User Manual](https://r-nimble.org/html_manual/cha-welcome-nimble.html) for more details.

## Dyes example

Let's pick a simple model from the classic WinBUGS examples: the dyes example. This is a simple normal hierarchical model. A description can be found [here](http://www.mrc-bsu.cam.ac.uk/wp-content/uploads/WinBUGS_Vol1.pdf).  A copy of the original is on our GitHub repository [here](https://github.com/nimble-dev/nimble/tree/devel/packages/nimble/inst/classic-bugs/vol1/dyes).  A modified version we will use is set up like this:
```{r}
library(nimble, warn.conflicts = FALSE)

dyesCode <- nimbleCode({
# Model
   for (i in 1:BATCHES) {
      for (j in 1:SAMPLES) {
         y[i,j] ~ dnorm(mu[i], sd = sigma.within);
      }
      mu[i] ~ dnorm(theta, sd = sigma.between);
   }
   
# Priors
   theta ~ dnorm(0.0, 1.0E-10);
   sigma.within ~ dunif(0, 100)
   sigma.between ~ dunif(0, 100)
})
```

Compared to the original, this has been modified by using standard deviation parameters instead of precision parameters in two places -- to illustrate NIMBLE's ability to handle different parameterizations -- and by removing the posterior predictive nodes.

By the way, any of the standard WinBUGS examples can be loaded automatically in NIMBLE like this:
```
classicDyesModel <- readBUGSmodel('dyes', dir = getBUGSexampleDir('dyes'))
```

## Create the model

We can create a model from `dyesCode` like this:
```{r}
dyesModel <- nimbleModel(dyesCode, constants = list(BATCHES = 6, SAMPLES = 5))
```

And we can set data values in it like this:
```{r}
data <- matrix(c(1545, 1540, 1595, 1445, 1595, 1520, 1440, 1555, 1550, 
1440, 1630, 1455, 1440, 1490, 1605, 1595, 1515, 1450, 1520, 1560, 
1510, 1465, 1635, 1480, 1580, 1495, 1560, 1545, 1625, 1445), nrow = 6)

dyesModel$setData(list(y = data))
dyesModel$y
```

It is also possible to set data values when calling `nimbleModel`.

## Use the model

Now the model is an R object, and we can manipulate it as such.

### Set or get values
```{r}
dyesModel$theta <- 1500
dyesModel$mu <- rnorm(6, 1500, 50)
dyesModel$sigma.within <- 20
dyesModel$sigma.between <- 20
dyesModel$y[1,]
dyesModel$theta
```

### Calculate log probability densities for part or all of the model
```{r}
## arbitrary example
dyesModel$calculate(c('theta', 'mu[1:6]', 'y[,2]'))
```

### Simulate part or all of the model
```{r}
## arbitrary example
dyesModel$mu
dyesModel$simulate(c('mu[1:3]'))
dyesModel$mu
```

### Query the model's relationships
```{r}
## arbitrary example
dyesModel$getDependencies(c('theta', 'mu[3]'))
```

### Plot the model graph (thanks to igraph's plot feature)
```{r}
library(igraph)
plot(dyesModel$getGraph())
```

With many nodes, the automatically generated graph can be hard to interpret. Note that arrows indicate node dependencies. The `mu` nodes depend on the centrally plotted `sigma.within`, `sigma.between`, and `theta` nodes, and the `y`s depend on their corresponding `mu`s. The purpose of this illustration is to visualize that NIMBLE models are represented as graphical objects that can be used programmatically.

## Compile the model

Finally we can compile the model and use the compiled version the same way we used the uncompiled version above:
```{r}
compiled_dyesModel <- compileNimble(dyesModel)
compiled_dyesModel$theta <- 1450
compiled_dyesModel$calculate() ## all nodes by default
```
Naturally, the compiled version is much faster.
