---
title: "Item Response Theory models"
output: html_document
bibliography: bibliography.bib
---

## Item Response Theory

Item Response Theory (IRT) refers to a family of models that attempt to explaining the relationship between latent traits and their manifestation in observations of the individual. The basic assumption is that an individual's observable behavior is related to this latent trait. Common contexts include educational testing and health status.

In a testing context, the tandard setting considers $I$ items and $P$ persons for which responses $y_{pi} \in \{0, 1\}$ are observed with values encoding incorrect and correct answers.
The probability of a correct answer for each item/person $\pi_{pi} = \text{Pr}(y_{pi} = 1)$ is modeled via logistic or probit regression, conditionally on a set of parameters encoding the characteristics of respondents and items. Typically a parameter $\eta_p$ is defined for each respondent $p = 1, \ldots, P$ representing the unobserved ability of the respondent. Parameters $\{\eta_p\}_1^P$ are assumed to be drawn from a common distribution, taking the role of 'random effects'. 

Different models can be defined according to the number of parameters associated with each item. Here we consider two classic models for dichotomous item responses, namely the one-parameter logistic (1PL) model and the two-parameter logistic (2PL) model. 

For this example we use data from the The Law School Admission Test (LSAT) which is classic example in educational testing [@DarrellBock1970], from the `ltm` package. The dataset comprises $1000$ responses to $5$ questions. 

```{r}
LSATdata <- ltm::LSAT
```

## 1PL model

The one-parameter logistic model (1PL) assumes the presence of one parameter, $\beta_i$, for each item $i = 1, \ldots, I$ encoding the *difficulty* of the item. We choose a non-informative normal prior $\mathcal{N}(0, 100)$ for the difficulty parameters, and a uniform distribution for abilities variance. 

```{r 1PLmodel}
library(nimble, warn.conflicts = FALSE)

code1PL <- nimbleCode({  
  for(i in 1:I) {
    for(p in 1:P) {
      y[p,i] ~ dbern(pi[p,i])
      logit(pi[p, i]) <-  eta[p] - beta[i]
    }
    beta[i] ~ dnorm(0, var = 100)
  }  
  
  for(p in 1:P) {
    eta[p] ~ dnorm(0, sd = sd_eta)
  }

  sd_eta ~ dunif(0, 100)  # prior for variance components based on Gelman (2006)
})

constants <- list(I = ncol(LSATdata), P = nrow(LSATdata))
data <- list(y = LSATdata)

set.seed(1)
inits <- list(beta    = rnorm(constants$I, 0, 1),
              eta     = rnorm(constants$P, 0, 1), 
              sd_eta = 3)

monitors = c("beta", "eta", "sd_eta")

model1PL <- nimbleModel(code1PL, constants, data, inits)
cModel1PL <- compileNimble(model1PL)
```

### Building and running an MCMC to fit the model

```{r 1PLmodel_compile}
conf1PL <- configureMCMC(model1PL, monitors = monitors)

model1PLMCMC <- buildMCMC(conf1PL)
cModel1PLMCMC <- compileNimble(model1PLMCMC, project = model1PL)
```

Run the model and save posterior samples. 

```{r 1PLmodel_run}
system.time(samples1PL <- runMCMC(cModel1PLMCMC, niter = 20000, nburnin = 10000))
```

### Checking the results

We can obtain a summary of the estimates of item difficulties.

```{r}
betaCols <- grep("beta", colnames(samples1PL))
sd_etaCol <- grep("sd_eta", colnames(samples1PL))
samplesSummary(samples1PL[, c(betaCols, sd_etaCol)])
```

Let's check posterior convergence. In this case, the variance component appears to be the slowest mixing, and
we would probably want to run the MCMC for longer.

```{r,  fig.cap='', fig.width=15, fig.height=5}
par(mfrow = c(1, 3), cex = 1.1)
for(i in 1:5)
  ts.plot(samples1PL[ , betaCols[i]], xlab = 'iteration', ylab = colnames(samples1PL)[ betaCols[i]])

ts.plot(samples1PL[ , sd_etaCol], xlab = 'iteration', ylab = colnames(samples1PL)[sd_etaCol])

```

### Item characteristic curve

We can use the posterior samples to estimate the posterior Item Characteristic Curve (ICC) for an interval of ability values. 

```{r 1PL_ICC}

## plot of ICC 
abilityGrid <- seq(-8, 5, length = 1000)
iccVals <- matrix(0, 1000, constants$I)
for(i in seq_len(constants$I)) {
	tmp <- sapply(abilityGrid, function(x) expit(x - samples1PL[, betaCols[i]]))
	iccVals[,i] <- colMeans(tmp)
}

library(ggplot2)
rownames(iccVals) <- abilityGrid
df <- reshape2::melt(iccVals)
colnames(df) <- c("Ability", "Item", "Probability")
df$Item <- factor(df$Item, levels = 1:5, labels = paste0("Item ", 1:5))

ggplot(df, aes(x = Ability, y = Probability, group = Item)) +
  geom_line(aes(color = Item)) + theme_bw() +
  ggtitle("Item characteristic curves - 1PL model")
```

```{r, echo = FALSE, eval = FALSE}

## r base version
# plot(abilityGrid, iccVals[,1], ylim = c(0,1), type = "l")
# lines(abilityGrid, iccVals[,2], ylim = c(0,1), type = "l", col = 2)
# lines(abilityGrid, iccVals[,3], ylim = c(0,1), type = "l", col = 4)
# lines(abilityGrid, iccVals[,4], ylim = c(0,1), type = "l", col = 3)
# lines(abilityGrid, iccVals[,5], ylim = c(0,1), type = "l", col = 5)

```

## 2PL model

A generalization of the 1PL model is the two-parameter logistic model, which considers an additional parameter, $\lambda_i$, for each item $i = 1, \ldots, I$, often referred as *discrimination* parameter, because items with a large $\lambda_i$ are better at discriminating between subjects with different abilities. Typically the $\lambda_i$'s are assumed to be positive, so we choose independent lognormal distributions as priors. 

Given the additional parameter, the metric (location and scale) of the individual parameters is only known up to a linear transformation. In the literature, several types of restrictions are proposed to anchor the metric, either acting on item or ability parameters. A simple restriction is to assume that abilities are drawn from a $\mathcal{N}(0,1)$ and freely estimate item parameters.


```{r 2PLmodel}
code2PL <- nimbleCode({
  for(i in 1:I) {
    for(p in 1:P) {
      y[p, i] ~ dbern(pi[p, i])
      logit(pi[p, i]) <-  lambda[i]*eta[p] + beta[i]
    }
  }  
  
  for(i in 1:I) {
    beta[i] ~ dnorm(0,  var = 100)  
    lambda[i] ~ dlnorm(meanlog = 0, sdlog = 6)
  }
   
  for(p in 1:P) {
    eta[p] ~ dnorm(0, 1)
  }
})

constants <- list(I = ncol(LSATdata), P = nrow(LSATdata))

data <- list(y = LSATdata)

set.seed(2)
inits <- list(beta   = rnorm(constants$I, 0, 3),
              eta    = rnorm(constants$P, 0, 3),
              lambda = rep(2, constants$I))

monitors = c("beta", "lambda", "eta")

model2PL <- nimbleModel(code2PL, constants, data, inits)

```

### Building and running an MCMC to fit the model

```{r 2PLmodel_compile}
cModel2PL <- compileNimble(model2PL)
conf2PL <- configureMCMC(model2PL, monitors = monitors)

model2PLMCMC <- buildMCMC(conf2PL)
cModel2PLMCMC <- compileNimble(model2PLMCMC, project = model2PL)

```


```{r 2PLmodel_run}
system.time(samples2PL <- runMCMC(cModel2PLMCMC, niter = 20000, nburnin = 10000))

```
### Checking the results

Similarly to the previous case, we can obtain posterior summaries and check the convergence and the item characteristic curves. Once again we would want to run the MCMC for longer. 

```{r 2PL_posterior_estimates}

betaCols <- grep("beta", colnames(samples2PL))
lambdaCols <- grep("lambda", colnames(samples2PL))

samplesSummary(samples2PL[, c(betaCols, lambdaCols)])
```

```{r 2PL_traceplots,  fig.cap='', fig.width=15, fig.height=5}
par(mfrow = c(1, 5), cex = 1.1)
for(i in 1:5)
  ts.plot(samples2PL[ , betaCols[i]], xlab = 'iteration', ylab = colnames(samples2PL)[ betaCols[i]])

par(mfrow = c(1, 5), cex = 1.1)
for(i in 1:5)
  ts.plot(samples2PL[ , lambdaCols[i]], xlab = 'iteration', ylab = colnames(samples2PL)[ lambdaCols[i]])

```

### Item characteristic curve


```{r 2PL_ICC}

abilityGrid <- seq(-8, 12, length = 1000)
iccVals <- matrix(0, 1000, constants$I)
for(i in seq_len(constants$I)) {
	## evaluate probability at each ability point
	tmp <- sapply(abilityGrid, function(x) expit(samples2PL[, lambdaCols[i]]*(x - samples2PL[, betaCols[i]])))
	iccVals[,i] <- apply(tmp, 2, mean)
}

rownames(iccVals) <- abilityGrid
df <- reshape2::melt(iccVals)
colnames(df) <- c("Ability", "Item", "Probability")
df$Item <- factor(df$Item, levels = 1:5, labels = paste0("Item ", 1:5))

ggplot(df, aes(x = Ability, y = Probability, group = Item)) +
  geom_line(aes(color = Item)) + theme_bw() +
  ggtitle("Item characteristic curves - 2PL model")
```

# References
