# Creating and running a Markov chain Monte Carlo algorithm in NIMBLE

This example shows how to quickly build and run a MCMC algorithm in NIMBLE with default sampler choices.

## pump example
Let's use another of the classic [WinBUGS](http://www.mrc-bsu.cam.ac.uk/software/bugs/) examples: the pump model (because it is really easy to show the model and the data).

A description can be found [here](http://www.mrc-bsu.cam.ac.uk/wp-content/uploads/WinBUGS_Vol1.pdf).  The original example can be found in our GitHub repository [here](https://github.com/nimble-dev/nimble/tree/devel/packages/nimble/inst/classic-bugs/vol1/pump).

### Create the model

We could load the model using `readBUGSmodel`, but instead we'll show it fully here:
```{r}
library(nimble)
pumpCode <- nimbleCode({ 
  for (i in 1:N){
      theta[i] ~ dgamma(alpha,beta)
      lambda[i] <- theta[i]*t[i]
      x[i] ~ dpois(lambda[i])
  }
  alpha ~ dexp(1.0)
  beta ~ dgamma(0.1,1.0)
})

pumpConsts <- list(N = 10,
                   t = c(94.3, 15.7, 62.9, 126, 5.24,
                       31.4, 1.05, 1.05, 2.1, 10.5))

pumpData <- list(x = c(5, 1, 5, 14, 3, 19, 1, 1, 4, 22))

pumpInits <- list(alpha = 1, beta = 1,
                  theta = rep(0.1, pumpConsts$N))

pump <- nimbleModel(code = pumpCode, name = 'pump', constants = pumpConsts,
                    data = pumpData, inits = pumpInits)
```

## Create the MCMC algorithm

```{r}
pumpMCMC <- buildMCMC(pump)
```

## Run the uncompiled MCMC if you want to

The model and algorithm can be used completely in R.  They will be really slow, but running in R allows easy testing and debugging.  I'm not going to actually run it for this document because it generates a bunch of warnings that are harmless but annoying.  Here is how you would run it for 10 iterations:
```
pumpMCMC$run(10)
```


## Compile the model and MCMC

It's more exciting to compile both the model and algorithm and run them.
```{r}
Cpump <- compileNimble(pump)
CpumpMCMC <- compileNimble(pumpMCMC, project = pump)
```

## Run it and look at the results
```{r}
CpumpMCMC$run(10000)
MCMCsamples <- as.matrix(CpumpMCMC$mvSamples)
plot(MCMCsamples[ , 'alpha'], type = 'l', xlab = 'iteration',  ylab = expression(alpha))
plot(MCMCsamples[ , 'beta'], type = 'l', xlab = 'iteration', ylab = expression(beta))
```