# Creating and running a Markov chain Monte Carlo (MCMC) algorithm in NIMBLE

This example shows how to quickly build and run a MCMC algorithm in NIMBLE with default sampler choices.

## Pump example
Let's use another of the classic [WinBUGS](http://www.mrc-bsu.cam.ac.uk/software/bugs/) examples: the pump model.

A description can be found [here](http://www.mrc-bsu.cam.ac.uk/wp-content/uploads/WinBUGS_Vol1.pdf).  The original example can be found in our GitHub repository [here](https://github.com/nimble-dev/nimble/tree/devel/packages/nimble/inst/classic-bugs/vol1/pump).

## Create the model

We could load the model using `readBUGSmodel()`, but instead we'll show it fully here:
```{r}
library(nimble)
pumpCode <- nimbleCode({ 
# Define relationships between nodes
  for (i in 1:N){
      theta[i] ~ dgamma(alpha,beta)
      lambda[i] <- theta[i]*t[i]
      x[i] ~ dpois(lambda[i])
  }
# Set priors
  alpha ~ dexp(1.0)
  beta ~ dgamma(0.1,1.0)
})

# Create some contrants, data, and initial values to pass to the model builder
pumpConsts <- list(N = 10,
                   t = c(94.3, 15.7, 62.9, 126, 5.24,
                       31.4, 1.05, 1.05, 2.1, 10.5))

pumpData <- list(x = c(5, 1, 5, 14, 3, 19, 1, 1, 4, 22))

pumpInits <- list(alpha = 1, beta = 1,
                  theta = rep(0.1, pumpConsts$N))
```

## Three ways to runMCMC

NIMBLE provides three different workflows for running MCMC giving you different levels of control over how the algorithm runs. From least to most control, these are: the `nimbleMCMC()` function, the `runMCMC()` function, and the `mcmc$run()` MCMC object function.

In general, we recommend the `runMCMC()` path, although `nimbleMCMC()` is useful for very straightforward cases.

### `nimbleMCMC()`

The `nimbleMCMC()` function handles everything from model building to compilation to running the MCMC. It takes model code and model inputs (constants, data, and initial values) and returns MCMC samples. The `nimbleMCMC()` function is easy and quick to use but provides little control over the MCMC algorithm. Also, since the objects (the uncompiled and compiled model and MCMC) are used only in the function, they aren't saved and can't be manipulated for debugging.


```{r}
nimbleMCMC_samples <- nimbleMCMC(code = pumpCode, 
                                 constants = pumpConsts, 
                                 data = pumpData, 
                                 inits = pumpInits,
                                 nburnin = 1000, niter = 10000)
head(nimbleMCMC_samples)
```

We can look at the results:

```{r}
plot(nimbleMCMC_samples[ , 'alpha'], type = 'l', xlab = 'iteration',  
     ylab = expression(alpha))
plot(nimbleMCMC_samples[ , 'beta'], type = 'l', xlab = 'iteration',
     ylab = expression(beta))
```

### `runMCMC()`

The `runMCMC()` function provides more control and a closer look at what NIMBLE is doing. `runMCMC()` takes a NIMBLE MCMC object, so we need to back up and build these things.

First, build the NIMBLE model.

```{r}
pumpModel <- nimbleModel(code = pumpCode, name = 'pump', constants = pumpConsts,
                    data = pumpData, inits = pumpInits)
```

Next, build an MCMC object for this model using `buildMCMC()`.

```{r}
pumpMCMC <- buildMCMC(pumpModel)
```

We can use `runMCMC()` on this uncompiled model if we want to, but we'll get a warning reminding us that compiled MCMC will be much faster. Uncompiled MCMC will be really slow, but running in R allows easy testing and debugging. Here is how you would run it for 5 iterations:

```{r}
runMCMC(pumpMCMC, niter = 5)
```

To compile the model and MCMC object, use `compileNimble()`. Note that the model object must be compiled first for the MCMC to compile.

```{r}
CpumpModel <- compileNimble(pumpModel)
CpumpMCMC <- compileNimble(pumpMCMC, project = pumpModel)
```

Now we can use `runMCMC()` on the compiled model object.

```{r}
runMCMC_samples <- runMCMC(CpumpMCMC, nburnin = 1000, niter = 10000)
```

We can look at the results:

```{r}
plot(runMCMC_samples[ , 'alpha'], type = 'l', xlab = 'iteration',  ylab = expression(alpha))
plot(runMCMC_samples[ , 'beta'], type = 'l', xlab = 'iteration', ylab = expression(beta))
```

### `mcmc$run()`

You can directly call the `$run()` method of an MCMC object. Similarly to `runMCMC()`, the MCMC object and model object must be built and compiled. Once this is done, you can use it as:

```{r}
CpumpMCMC$run(10000)
run_samples <- as.matrix(CpumpMCMC$mvSamples)

plot(runMCMC_samples[ , 'alpha'], type = 'l', xlab = 'iteration',  ylab = expression(alpha))
plot(runMCMC_samples[ , 'beta'], type = 'l', xlab = 'iteration', ylab = expression(beta))
```

There are several advantages to using `runMCMC()` over `$run()`. `runMCMC()` can handle multiple chains and burn-in. It also returns a friendlier output, whereas `$run()` requires retrieving samples from the MCMC object and converting to a matrix.



